---
title: "GIS Exploratory Data Analysis"
author: "Group P4T2"
date: "2024-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(tidyverse)
library(sf)
library(dplyr)
library(tidyr)
```

# Data Data Cleaning for GIS Analysis

```{r read data}

# Read the PSE and Equity shape files
pse_data <- st_read("../Data/PSE_Gas_Tracts_2.shp")
equity_data <- st_read("../Data/Race_and_Social_Equity_Composite_Index_for_2020_Census_Tract_Geographies.shp")

```

pse_data aggregates gas consumption data by sector, year (2020-2023), and census tract in Seattle. Equity_data integrates race, ethnicity, socioeconomic status, and health disadvantages in Seattle.

Below we will create three data frames. 
The first data frame (GIS_EE_2023) will only include pse_data from 2023 and be merged with the equity data. 
The second data frame (GIS_Equity) will be include all of the equity_data. 
Finally, the third data frame (GIS_Energy), will only contain pse_data.

For equity_data we will merge on GEOID and pse_data will merge on Census_Tract. 

```{r column names}
# Get column names of pse_data and equity_data
colnames(pse_data)
colnames(equity_data)
```


```{r GIS 2023 - Merge Data}

# Step 1: Filter the PSE data for the year 2023
pse_data_2023 <- subset(pse_data, CAL_YEAR == 2023)

# Step 2: Ensure the GEOID and Census_Tract columns are of the same type before merging
pse_data_2023$Census_Tra <- as.character(pse_data_2023$Census_Tra)
equity_data$GEOID <- as.character(equity_data$GEOID)

# Step 3: Convert to regular dataframes for the merge operation, and drop geometry column from equity_data
pse_data_df <- as.data.frame(pse_data_2023)
equity_data_df <- as.data.frame(equity_data %>% st_set_geometry(NULL))

# Step 4: Merging the filtered 2023 PSE data with the equity data on GEOID/Census_Tract
merged_data_df <- inner_join(pse_data_df, equity_data_df, by = c("Census_Tra" = "GEOID"))

# View the merged dataframe
head(merged_data_df)

# Step 5: Reattach the geometry to merged_data_df
# Since merged_data_df is no longer an sf object after the join, convert it back
# Prepare a geometry column from pse_data_2023 matching Census_Tra in merged_data_df
geometry_col <- pse_data_2023 %>% 
  select(Census_Tra, geometry) %>% 
  distinct(Census_Tra, .keep_all = TRUE)

# Join this geometry data back to the merged dataframe
GIS_EE_2023 <- merged_data_df %>%
  left_join(geometry_col, by = "Census_Tra") %>%
  st_as_sf(crs = st_crs(pse_data_2023))

# View the final sf object
head(GIS_EE_2023)


```


This code filters pse_data for entries from the year 2023 and ensures that the identifiers (Census_Tra from pse_data_2023 and GEOID from equity_data) are of compatible data types for merging. It then converts both datasets to regular dataframes, explicitly removing the geometry column from equity_data to simplify the merge operation. 

An inner_join is performed based on these identifiers, merging attributes from equity_data into pse_data_2023 where their identifiers match. Finally, the merged dataframe is converted back into an sf object by reattaching the geometry from pse_data_2023. 

This process ensures a streamlined dataset that combines relevant attributes from both sources, based on the year 2023 and matching geographic identifiers, for detailed spatial analysis.


```{r GIS 2023 - Clean Data}

```



```{r GIS Equity Data}

# View the Racial and Social Equity data
head(equity_data)

```




```{r GIS Energy}

# View the PSE (Puget Sound Energy) Usage
head(pse_data)

```



