---
title: "Exploratory Data Analysis for Seattle Building Energy Analysis"
author: "P4T2"
date: "2024-03-05"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(sf)
```

# Introduction

Briefly describe the project, objectives, and data sources. Test, Connor working w git.

```{r read data}

# Read the PSE and Equity shape files for GIS
pse_data <- st_read("../Data/PSE_Gas_Tracts_2.shp")
equity_data <- st_read("../Data/Race_and_Social_Equity_Composite_Index_for_2020_Census_Tract_Geographies.shp")

```

# Data Cleaning

```{r Identifying Missing Values}



```

Discuss your strategy for identifying and handling missing values.

```{r Identifying Outliers}



```

Strategy and code for identifying and handling outliers.

```{r Special Coding for Numerical Data (Binning, etc.)}



```

```{r Converting and Manipulating Character Data}



```

```{r Ensuring Data Integrity}



```

Discuss data integrity, type checking, format, bounds.

```{r}

```

# Data Reshaping

Discuss if and why your dataset needed reshaping.

# Data Management

Apply data management strategies, summarize key variables, and address outliers.

In our project, we've implemented a streamlined collaborative workflow on GitLab, focusing on minimizing collaboration errors through effective data management strategies. Each team member initiates their work in a dedicated branch. This allows for isolated development without impacting the main codebase. Regular commits are made within these branches using `git add .` and `git commit -m "description"`.

Before we start any work in our individual branch, first we update our branch with the main branch, using `git pull origin main`. This streamlines our development process, and also enforces a disciplined data management practice that minimizes errors and fosters efficient collaboration among team members.

# Exploratory Data Analysis

```{r Plot Distributions of Key Variables}



```

```{r Key Metrics/Measurements/Statistics}



```

```{r Analysis of Outliers/Anomalies}



```

```{r Deploying Graphical Tools}

```

```{r Analyzing Data Over Time}

```

```{r Comparison Across Categories}

```

```{r Pattern Identification and Interpretation}

```

# Discussion and Conclusions

Summarize the key findings and outline any recommendations.

# References

List all data sources and other references used in your analysis.
